Supporting Information:\\Black-swan events in animal populations

### 2016-06-23 text/ms.tex
Add my edits to manuscript before Nick/Trevor edits

Sean C. Anderson\textsuperscript{1,2*}, Trevor A.
Branch\textsuperscript{3}, Andrew B. Cooper\textsuperscript{2}, Nicholas K.
Dulvy\textsuperscript{1}

\textsuperscript{1}Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada

\textsuperscript{2}School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada

\textsuperscript{3}School of Aquatic and Fishery Sciences, Box 355020,
University of Washington, Seattle, WA 98195, USA

\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail:
sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box
355020, University of Washington, Seattle, WA 98195, USA

@@

Keywords:
Bayesian,
die-offs,
ecological surprises,
Gompertz,
heavy tails,
leptokurtic,
mass mortality,
population dynamics,
risk,
Student t

Words in abstract: TOdO

Words in main text: todo

@@

Number of figures: 4

Number of tables: 0

@@

Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as
population die-offs and crashes (\percBSDown \%) driven by extrinsic
perturbations from climate variability. Thus, ignoring black-swan events will
underestimate extinction risk. Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to future increases in climate
extremes.

@@

Indeed, such die-offs may be the most important element
affecting population persistence \citep{mangel1994} and their importance is
likely to increase given projected increases in the frequency and magnitude of
climate-related extremes \citep{ipcc2012}.
% Despite this anecdotal evidence for
% ecological black swans, systematic searches of large numbers of timeseries have
% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
However, the overwhelming majority of population model fitting and risk forecasting
assumes that deviations from model predictions can be represented by a normal
distribution \citep[typically on a log scale; e.g.][]{brook2006a, knape2012}.
If black swans occur, though, a normal distribution would under-estimate the
probability of extreme events occurring.

There are many examples of black-swan events in populations
but we lack a large-scale comparative analysis
with a method that allows for
population dynamics
(e.g.\ per capita productivity, density dependence, and typical stochasticity).
Previous comparative analyses fitting
heavy-tailed distributions to timeseries
(distributions with higher probabilities of extreme events
than the normal distribution)
have not accounted for an underlying population dynamics model
\citep{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
jumps in abundance that exceed a threshold of decline
\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
But this approach gives no indication of relative frequency
and again does not distinguish events from
expected dynamics.
Methodologically, the generalized extreme value (GEV) distribution represents
the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
because the GEV requires maxima per time block
or peaks beyond a chosen threshold,
the GEV requires longer timeseries then typically available
for populations and is not simply integrated
into a population dynamics model.

Here, we develop a new approach to estimate the frequency and magnitude of
black-swan dynamics across time series of 609 populations from a wide array of
taxonomic groups---including many  birds, mammals, insects, and fishes
(Table~S1). We identify characteristics of time series or
life-history traits associated with the detection of black-swan events and
verify known causes. To accomplish this, we develop a framework for identifying
heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether
the largest stochastic jumps in log abundance from one time step to the next
are more extreme than typically seen with a normal distribution. Our framework
allows for a range of population dynamic models, can incorporate observation
uncertainty and skewness in process noise, and can be readily applied to
abundance time series.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-t distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1,
Fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed
distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a t distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

We picked the t distribution mainly for pragmatic reasons:
it requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations are not present.
In addition, the t distribution has some mechanistic justification.
The t distribution can be formed by a mixture of normal distributions
with the same mean and different variances \citep{gelman2014}.
This could represent the effect of unmodeled spatial heterogeneity,
a nonstationary environmental process, or
varying responses to temperature among individuals,
for example.

@@

We selected abundance time series from the Global Population Dynamics
Database \citep{gpdd2010} (GPDD), which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated some missing values (sensu \citet{brook2006a}). Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD. The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
\end{align}
The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-t distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

For the Gompertz model, our weakly-informative priors (Fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our Half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain
convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by \citet{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid drifting towards infinity. In
any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the t distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
population as heavy tailed.

We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by \citet{gelman2006a} of
\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
confidence intervals. In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains were deemed to have sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are mostly concerned about when categorizing
populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here (not shown).

\subsection{Alternative population models} We fit alternative population models to
test if four key phenomena systematically changed our conclusions. The range of
percentages of black swans by taxonomic class cited in the abstract are based
on lower and upper limits across our main Gompertz model and these four
alternative models.

@@

\subsubsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep{valpine2002, thorson2014b}.

@@

We found that we could consistently and precisely recover median posterior
estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
200\)) (Fig.~\ref{fig:sim-nu} upper panels). At smaller samples we could
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~\ref{fig:sim-nu} lower panels).

Our simulation-testing results indicate that the Gompertz model can recapture
the true value of \(\nu\) when the process noise was chosen so that appropriate
tail events were present (Fig.~\ref{fig:sim-prob} upper panels). The addition
of observation error caused the model to tend to underestimate the degree of
heavy-tailedness. Fitting a model with correctly specified observation error
did not make substantial improvements to model bias (Fig.~\ref{fig:sim-prob}).

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from \citet{brook2006a}.
The beta distribution is useful when response data are on a continuous scale
between zero and one \citep{ferrari2004}. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + X_i \beta),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(X_i\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\beta\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to match the
way they are visually represented in Fig.~\ref{fig:correlates} and to make the
relationship approximately linear on the logit-transformed response scale. All
input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in
magnitude \citep{gelman2008c}. We excluded body length as a covariate because it
was highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.

>>>>

Sean C. Anderson\textsuperscript{1*}, Trevor A.
Branch\textsuperscript{2}, Andrew B. Cooper\textsuperscript{3}, Nicholas K.
Dulvy\textsuperscript{4}

\textsuperscript{1}Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
sandrsn@uw.edu

\textsuperscript{2}School of Aquatic and Fishery Sciences, Box 355020,
University of Washington, Seattle, WA 98195, USA, e-mail:
tbranch@uw.edu

\textsuperscript{3}School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada, e-mail:
andrew\_cooper@sfu.ca

\textsuperscript{4}Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
nick\_dulvy@sfu.ca

\textsuperscript{*}Corresponding author; present address:
School of Aquatic and Fishery Sciences, Box
355020, University of Washington, Seattle, WA 98195, USA

@@

Keywords:
Bayesian,
die-offs,
ecological surprises,
Gompertz,
heavy tails,
leptokurtic,
mass mortality,
population dynamics,
risk,
Student-t distribution

Words in abstract: 142

Words in main text: 4820

@@

Number of figures/tables: 4/0

Statement of authorship:
S.C.A. and T.A.B. conceived the project;
S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
data and wrote the paper with input from all authors.

@@

Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
and manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to such extremes.

@@

Indeed, such die-offs may be the most important element
affecting population persistence \citep{mangel1994} and their importance is
likely to increase given projected increases in the frequency and magnitude of
climate-related extremes \citep{ipcc2012}.
% Despite this anecdotal evidence for
% ecological black swans, systematic searches of large numbers of timeseries have
% yet to reveal black-swan events \citep{keitt1998, allen2001, halley2002}.
However, the overwhelming majority of population model fitting and risk forecasting
assumes that deviations from model predictions can be represented by a normal
distribution \citep[on a log scale; e.g.][]{brook2006a, knape2012}.
If black swans occur, though, a normal distribution would under-estimate the
probability of extreme events occurring.

There are many examples of black-swan events in populations
but we lack a large-scale comparative analysis
with a method that allows for
population dynamics
(e.g.\ per capita productivity, density dependence, and typical stochasticity).
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to timeseries
have not accounted for an underlying population dynamics model
\citep{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
jumps in abundance that exceed a threshold of decline
\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
But this approach gives no indication of relative event frequency
and again does not distinguish events from
expected dynamics.
Methodologically, the generalized extreme value (GEV) distribution represents
the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
because the GEV requires maxima per time block
or peaks beyond a chosen threshold,
the GEV requires longer timeseries then typically available
for wild populations and is not simply integrated
into a population dynamics model.

Here, we develop a new approach to estimate the frequency and magnitude of
black-swan dynamics across time series of 609 populations from a wide array of
taxonomic groups---including many  birds, mammals, insects, and fishes
(Table~S1). We identify characteristics of time series or
life-history traits associated with the detection of black-swan events and
verify known causes. To accomplish this, we develop a framework for identifying
heavy-tailed (black-swan) process noise in population dynamics.
We test if the largest stochastic jumps in log abundance from one time step to the next
are more extreme than typically seen with a normal distribution. Our framework
allows for a range of population dynamic models, can incorporate observation
uncertainty and skewness in process noise, and can be readily applied to
abundance time series.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-t distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a t distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

We picked the t distribution for pragmatic and mechanistic reasons.
Pragmatically, the t distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations are not present.
Mechanistically, the t distribution can be formed from a mixture of normal distributions
with the same mean and different variances \citep{gelman2014}.
This could represent the effect of unmodeled spatial heterogeneity,
nonstationary environmental processes, or
varying responses to temperature among individuals, for example.

@@

We selected abundance time series from the Global Population Dynamics
Database \citep{gpdd2010} (GPDD), which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated remaining missing values (sensu \citet{brook2006a}).
Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD. The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the ln abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-t distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

For the Gompertz model, our weakly-informative priors (Fig.~S\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\ \lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\ \nu &\sim
\mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our Half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep[e.g.][]{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain
convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by \citet{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid drifting towards infinity. In
any case, for \(\nu > 20\) the t distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the t distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) is unlikely to flag the
population as heavy tailed.

We fit our models with Stan 2.6.0 \citep{stan-manual2015, hoffman2014}, and R
3.2.0 \citep{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by \citet{gelman2006a} of
\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
confidence intervals. In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains had sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \citep{stan-manual2015}.

@@

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~S\ref{fig:priors}c). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are mostly concerned about when categorizing
populations as heavy tailed. We furthermore considered a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here (not shown).

\subsection{Alternative population models} We fit alternative population models to
test if four key phenomena systematically changed our conclusions.

@@

\subsubsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the unobserved state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

@@

\textit{TODO: Should the following 2 paragraphs go in the results?}
We found that we could consistently and precisely recover median posterior
estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~S\ref{fig:sim-nu} lower panels).

Our simulation-testing results indicate that the Gompertz model can recapture
the true value of \(\nu\) when the process noise was chosen so that appropriate
tail events were present (Fig.~S\ref{fig:sim-prob} upper panels). The addition
of observation error caused the model to tend to underestimate the degree of
heavy-tailedness. Fitting a model with correctly specified observation error
did not make substantial improvements to model bias (Fig.~S\ref{fig:sim-prob}).

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from \citet{brook2006a}.
The beta distribution is useful when response data are on a continuous scale
between zero and one \citep{ferrari2004}. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \bm{X_i} \bm{\beta}),
  \: \text{for } i = 1, \dots, 617\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to match the
way they are visually represented in Fig.~S\ref{fig:correlates} and to make the
relationship approximately linear on the logit-transformed response scale. All
input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in
magnitude \citep{gelman2008c}. We excluded body length as a covariate because it
was highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.
